**Title**: Linking pinyon jay population abundance to pinyon cone production and climate in the SW United States

### 2024-09-24 drafts/draftAug2024.qmd
updated code to buffer around state boundaries and used agreed upon cone data (z-score abundances) and updated basal area data from Kyle. running on monsson

In the SW, there is a species of jay that has been declining over the last several decades. This jay is co-evolved with pinyon pine trees, whose seeds it caches across the landscape to eat later or to forget later and help the tree disperse to new areas. Pinyon pines have also seen dramatic shifts in range (expansions, die offs) and reproduction.

Given the complex co-evolution of the interaction between the pinyon pine and the pinyon jay, it is hard to tease apart the cause and effect of these patterns or how changes in resource availability alter jay abundances. This has not been evaluated to date. It is likely that, given the broad geographical range of both species, that a different combination of forces may be playing out across the species' range, and that these effects may not be concurrent. Pinyon pine cone production occurs \~2 years following favorable climate years. Other food resources for jays in these systems may respond more quickly to the same climate conditions (e.g., insects, junipers, and gambel oak). Thus, it may be that jay populations increase \*prior\* to good cone years for pinyon pines as they are either signaled to "get ready" by the same climate conditions as trees or because their populations get a boost prior to a good cone year from other food resources. The result of jays "predicting" good cone years is that more jays will be in the landscape to distribute seeds to more new places. In addition or conversely, jays could also respond with an immediate or lagged effect to cone production. Teasing these patterns out and how other climatic conditions mediate these effects could be imperative for understanding the link between patterns in resources and birds and help in conservation of the species.

On a more methodological note, further muddying the waters is the fact that while we have several broad-scale monitoring programs that capture pinyon jays, they all have limitations both spatially and temporally. Ideally, we would like to combine them into one model that helps us predict jay-pine relationships.

In this study, we pulled two data sources for pinyon jays (BBS and eBIRD) into the same statistical model and combining them into a joint likelihood -- where both datasets contribute to estimates of population numbers. (this method is called "data integration"). We then used SAM models to explore the timescales at which jays either predict or respond to pinyon pine cone production in the context of a suite of other broad-scale climatic mediates (temperature, precip, monsoonality).

@@

eBIRD data is unstructured data without a regular sampling design, so we had to spatially sub-sample eBIRD data to limit spatial bias. To do this, we implemented a modified version of a previously-used method that links checklists with and without observations of a species of interest within a certain spatial scale as repeat measurements at that spatial scale (e.g., raster grid cell) [@schindler2022]. We adapted the method in that study; however, to not explicitly select checklists within specific raster grid cells. Rather, we linked observed-not observed checklists based on spatial proximity (1km distance or less). We then spatially subset these pairs along with all checklists (observed and not observed) such that within a given year, all checklists (or checklist pairs) were at least 4km from another checklist (or pair).

@@

(ALSO include methods on accounting for spatial uncertainty in bird observation datasets)

@@

\[explanations of the biological process model and of SAM component\]

| Covariate or interaction           | Explanation                                                                                                                |
|------------------------------------|----------------------------------------------------------------------------------------------------------------------------|
| Pinyon pine cone abundance         | Modeled abundance of cone availability (can have lead and lagged effects)                                                  |
| Maximum temperature                | A proxy for physiological limits and/or insect resource availability (can have lagged effects)                             |
| Cumulative precipitation           | A proxy for insect resource availability (can have lagged effects)                                                         |
| Monsoonality                       | important in conjunction with cone interaction (see below for interactions)                                                |
| Pinyon basal area                  | important in conjunction with cone interaction (see below for interactions)                                                |
| Cone abundance x Temperature       | Is the importance of cone abundance for jays mediated by temperature?                                                      |
| Cone abundance x Precipiation      | Is the importance of cone abundance for jays mediated by precipitation (e.g., more/less availability of insect prey)       |
| Cone abundance x Monsoonality      | Is the importance of cone abundance for jays different across a range of monsoonal conditions that alter cone reliability? |
| Cone abundance x Pinyon basal area | A combined metric of actual seed availability - is jay abundance linked to availability of cone abundance on the landscape |

@@

>>>>

\[big picture\]

Co-evolved relationships and how they may be changing with climate change. Additionally, how we may draw understanding about the diversity of responses because these co-evolved relationships occur on the background of other broad-scale variation in species' ranges that might mediate (good or bad, helping/hurting) the co-evolution and its response to global change drivers.

\[zooming in\]

In the SW, there is an important co-evolution between pinyon pine and pinyon jays. Jays have been declining over the last several decades. This jay caches pinyon seeds across the landscape to eat or to forget later and help the tree disperse to new areas. Pinyon pines have also seen dramatic shifts in range (expansions, die offs) and reproduction due to climate change and climate variability. \[Some biology about pinyon masting. Some more info on the nature of the co-evolution here too\]

\[gap\]

\[What is causing their decline?\] Given the complex co-evolution of the interaction between the pinyon pine and the pinyon jay, it is hard to tease apart the cause and effect of these patterns or how changes in resource availability alter jay abundances. This has not been evaluated to date. It is likely that, given the broad geographical range of both species, that a different combination of forces may be playing out across the species' range, and that these effects may not be concurrent. Pinyon pine cone production occurs \~2 years following favorable climate years. Other food resources for jays in these systems may respond more quickly to the same climate conditions (e.g., insects, junipers, and gambel oak). Thus, it may be that jay populations increase \*prior\* to good cone years for pinyon pines as they are either signaled to "get ready" by the same climate conditions as trees or because their populations get a boost prior to a good cone year from other food resources. The result of jays "predicting" good cone years is that more jays will be in the landscape to distribute seeds to more new places. In addition or conversely, jays could also respond with an immediate or lagged effect to cone production. Teasing these patterns out and how other climatic conditions mediate these effects could be imperative for understanding the link between patterns in resources and birds and help in conservation of the species.

\[Additional challenge of data availability at scales that match covariates.\] On a more methodological note, further muddying the waters is the fact that while we have several broad-scale monitoring programs that capture pinyon jays, they all have limitations both spatially and temporally. Ideally, we would like to combine them into one model that helps us predict jay-pine relationships.

\[our study\]

In this study, we aimed to determine the relationship between pinyon pine cones and pinyon jay abundance in the Southwest US and how other site-level factors may mediate this relationship. To leverage the broadest spatial and temporal availability of data on pinyon jay abundance, we used a joint-likelihood model that combined observed counts of pinyon jays from Breeding Bird Survey (BBS, cite) and eBIRD (Cite) data into a single model estimating latent abundance (CITE). We then evaluated how pinyon jay abundance related to cone production (CITE andreas), pinyon pine basal area, climate (maximum temperature, precipitation, and monsoonality), and the interactions between cone production and these other variables as broad-scale mediators of the relationship between cone availability and alternative food sources in a given site and a given year (e.g., seeds, acorns, insects). We allowed cone availability to have immediate, lead (jays "predict" cone years), and lag (jays "respond to" cone years) effects on jay abundance using a stochastic antecedent modeling approach (CITE). Relatedly, we allowed maximum temperature and precipitation to have lagged and immediate effects on jay abundance. Specifically, in this study, we wanted to evaluate 1) how strong is the relationship between pinyon jay abundance and cone availability? and 2) how is this relationship mediated by habitat and climate variables that vary by site and/or year? This study provides a valuable way of determining the strength of co-evolution between pinyon pines and pinyon jays and valuable insights into the broad-scale mechanisms by which pines and jays may be declining in the Southwest US and what might be mediating this decline.

@@

eBIRD data is unstructured data without a regular sampling design, so we had to spatially sub-sample eBIRD data to limit spatial bias. To do this, we implemented a modified version of a previously-used method that links checklists with and without observations of a species of interest within a certain spatial scale as repeat measurements at that spatial scale (e.g., raster grid cell) [@schindler2022]. We adapted the method in that study; however, to not explicitly select checklists within specific raster grid cells. Rather, we linked observed-not observed checklists based on spatial proximity (1km distance or less). We then spatially subset these pairs along with all checklists (observed and not observed) such that within a given year, all checklists (or checklist pairs) were at least 4km from another checklist (or pair). The coordinates for a pair became their midpoint; coordinates for single observed-not observed checklists remained the same.

@@

In a joint-likelihood model, multiple independent datasets contribute to a latent "true" value for a variable. In this model, we are combining observed counts from BBS and eBIRD to estimate a "true" latent abundance at each grid, $i$, in each year $y$, $N_{i,y}$, of pinyon jays in a given 4x4 km grid cell in a given year. Both of these observation datasets are part of separate observation process models that incorporate covariates that impact the detection of birds within each survey. For both observation processes, observed count data are binomially distributed based on a detection probability for a site in a year and the "true" latent abundance, $N_{i,y}$, being the total possible "trials".

BBS surveys are conducted on a set of transects. In this study, we have subset the first 10 stops on any transect - these stops are considered spatial replicates for that transect and all contribute to one estimate of detection probability for that transect. We considered observer experience (in years) as a covariate to detection of pinyon jays for BBS surveys. The transects are located in a 4x4 km grid cell and transect-level information is linked back to this grid cell based on $N_{i,y}$ (more information about the process for this below \_\_\_). BBS count data for each transect, $t$, year, $y$, and stop, $r$, is distributed as:

$$
BBSCount_{t,y,r} \sim Binomial(N_{i(t),y}, BBSDetectionProb_{t,y})
$$

The detection probability for BBS data includes a regression with the detection covariate of observer experience (in years), which depends on site and year, but not on stop (the same observer conducted all stops on a transect in a year):

$$
logit(BBSDetectionProb_{t,y}) = \beta_0 + \beta_1*ObserverExperience_{t,y}
$$

eBIRD surveys are conducted at a set of points with a set survey distance. We spatially sub-sampled eBIRD observations (described above) so that pairs of observed-absent checklists or single checklists of either type were 4 km from each other. We buffered checklists so that none occurred within 10km of the "edge" of the four states for which we had cone abundance data (Utah, Colorado, New Mexico, and Arizona). Again, eBIRD observed count data for pair, $g$, in year, $y$, in checklist, $c$, are binomially distributed based on a detection probability and the latent true abundance of birds in the 4x4 km grid cell in which the checklist is located:

$$
eBIRDCount_{g,y,c} \sim Binomial(N_{i(g),y}, eBIRDDetectionProb_{g,y,c})
$$

The detection probability for eBIRD data includes a regression with the detection covariates of survey type (stationary or traveling), start time, duration of survey, distance traveled, and the number of observers for each checklist. See above (\_\_\_) for how we filtered these variables initially based on best practices so that checklists represented similar effort. These covariates all depended on group, year, and checklist:

$$
logit(eBIRDDetectionProb_{g,y,c}) = \gamma_0 + \sum_{j=1}^{5}\gamma_j*X_{j_{g,y,c}}
$$

For both datasets, we wanted to account for spatial uncertainty in both the coordinates provided in the datasets as well as where within each survey pinyon jays were actually detected (e.g., we did not know if jays were detected at the beginning of the survey or halfway into the survey, or the end of a survey - a distance that could be up to 8km away from the starting location). To account for this spatial uncertainty, we determined which 4x4 km grid cell each survey (BBS transect or eBIRD checklist pair) was located in based on a stochastic process that depended on what proportion of a buffered radius around each survey was contained within a given grid cell. We sampled from possible grid cells using these proportions as a vector of probabilities for each transect or pair and year, $\pi_{t/g, y}$, of birds being located in that grid cell based on a categorical distribution. Thus, for each iteration, $iter$, of the model, the grid cell ID for either eBIRD or BBS is given as:

$$
gridID_{iter,t/g, y} \sim Categorical(\pi_{t/g, y})
$$

@@

As highlighted in the section above, the joint observation processes contribute to a shared estimate of "true" latent abundance, $N_{i,y}$, of pinyon jays in a given 4x4 km grid cell in a given year. $N_{i,y}$ is Poisson distributed around an expected rate parameter, $\lambda_{i,y}$, as follows:

$$
N_{i,y} \sim Poisson(\lambda_{i,y})
$$

Furthermore, $\lambda_{i,y}$ is dependent on a regression of environmental covariates:

$$
log(\lambda_{i,y}) = \alpha_0 + \sum_{j=1}^J \alpha_j*X_{j_{i,y}} + \sum_{k=1}^K \alpha_k*Z_{k_{i,y}} + \sum_{m=1}^4\sum_{n=m+1}^4 \alpha_{mn}*X_{m_{i,y}}*X_{n_{i,y}}
$$

In this formulation, covariates $X_{j_{i,y}}$ include only concurrent effects. These variables in the model include a cell-level monsoonality score and pinyon pine basal area (yearly). Conversely, values for each $Z_{k_{i,y}}$ comprise a stochastic antecedent structure [@ogle2015] dependent on the covariate, where each $Z_{k_{i,y}}$ is the weighted average value of a set of seasonal or yearly values for that covariate across a defined set of seasons or years. The weights for each covariate across all seasons or years, *l*, $w_{k,l}$, have a Dirichlet prior and sum to 1. The weight, or importance weight, for each season or year indicates how important that value is to the overall effect (larger values of the weight mean more importance of that season or year to that covariate effect). The covariates for which antecedent weights are calculated are cone production, maximum temperature, and precipitation. Cone production is considered on a yearly scale (cones only produce up to one time in a year) and include both lead and lagged effects (e.g., jays "predict" cone years or "respond" to good cone years), and include \_\_\_ years prior years, the current year, and \_\_ years after the current year. Both temperature and precipitation are at a seasonal scale and only consider lagged effects. The seasons are based on important seasonal biology for pinyon jays, including: breeding (February-April), feeding dependent young (May-June), summer (July), fall and winter foraging, potentially with irruption (August - January) [@wigginsd.a.2005]. For seasonal covariates, we considered \_\_ seasons into the past. We also considered interactions between cone production and all other covariates (pinyon basal area, monsoonality, maximum temperature, and precipitation) to determine how the relationship between cones and birds is mediated by other covariates related to cone reliability and alternative food resources (e.g., juniper berries, acorns, insects).

| Covariate or interaction           | Description (all are at 4x4 km grid cell scale)                                                | Question                                                                                                                          |
|------------------|---------------------------|---------------------------|
| Pinyon pine cone abundance         | Abundance of cones from a predictive model (CITE)                                              | How is pinyon jay abundance linked to pinyon cone availability (can have lead, immediate, and lagged effects)?                    |
| Maximum temperature                | A proxy for physiological limits and/or insect resource availability (can have lagged effects) | How is pinyon jay abundance linked to physiological limits and/or insect availability?                                            |
| Cumulative precipitation           | A proxy for insect resource availability (can have lagged effects)                             | How is pinyon jay abundance linked to insect availability?                                                                        |
| Monsoonality                       | The degree to which the site depends on monsoon moisture                                       | How is pinyon jay abundance shaped by broad-scale climatic mediators of cone reliability?                                         |
| Pinyon basal area                  | The basal area of pinyon pine                                                                  | How is pinyon jay abundance shaped by local habitat preferences for food trees?                                                   |
| Cone abundance x Temperature       |                                                                                                | Is the importance of cone abundance for jays mediated by temperature?                                                             |
| Cone abundance x Precipiation      |                                                                                                | Is the importance of cone abundance for jays mediated by precipitation (e.g., more/less availability of insect prey)              |
| Cone abundance x Monsoonality      |                                                                                                | Is the importance of cone abundance for jays different across a range of monsoonal conditions that alter cone reliability?        |
| Cone abundance x Pinyon basal area |                                                                                                | Is the importance of cone abundance for jays different depending on the local habitat configuration and abundance of pinyon pine? |

@@

We implemented models in the Bayesian software JAGS (CITE) using R (CITE) and the jagsUI wrapper package (CITE). We prepared data using the here (CITE), tidyverse (CITE), sf (CITE), terra (CITE), readxl (CITE), sf (CITE), exactextractr (CITE), spatialEco (CITE), nngeo (CITE), sp (CITE), auk (CITE), lubridate (CITE), prism (CITE), and data.table (CITE) packages.

We initially ran models to generate 4000 MCMC samples with which to determine how many iterations would be needed to reach convergence (CITE). We then re-ran models with more iterations and initial values sufficient to reach convergence based on the raftery.diag() function in the coda package (CITE). We assessed model convergence by evaluating trace, history, and autocorrelation plots generated with the mcmcplots package (). We quantified model convergence by evaluating whether Gelman-Ruman statistics, $\hat{R}$, for all root nodes in the model were \< 1.2. We assessed model goodness-of-fit by comparing observed count data from eBIRD and BBS to predicted count data for each dataset based on the model mean value.
